Apparent preferences for cognitive effort fade when multiple forms of effort and delay are interleaved in a foraging environment

### 2020-10-29 Cost_manuscript.Rmd
draft for Joe number 1

  Evaluating whether to pursue a desired outcome often requires assessing the demands required for obtaining it. In psychology and economics, demands for both time and effort have traditionally been identified as costs that detract from the net subjective value of rewards (Westbrook & Braver, 2015). Indeed, decision makers tend to prefer low-effort and immediate rewards over larger rewards that demand longer delays and higher effort (Ainslie, 1975; Hull, 1943; Kool & Botvinick, 2018; Kool, McGuire, Rosen, & Botvinick, 2010; Treadway, Buckholtz, Schwartzman, Lambert, & Zald, 2009; Westbrook, Kester, & Braver, 2013). However, people often change their attitudes towards exerting effort in everyday decisions, resulting in pursual of effortful avenues instead. Emerging experimental and theoretical work has chiefly explained this phenomenon by suggesting a reciprocal relationship between effort and outcome, whereby exerting effort boosts the perceived value of its outcomes (Kacelnik and Marsh, 2002; Mochon, Norton, and Ariely, 2012; Hernandez Lallement et al., 2014), and positive outcomes imbue value into their preceding effortful behavior (i.e. learned industriousness, Eisenberger, 1992). Though intuitively valid and empirically supported, this explanation leaves open questions about what intrinsic features of an effortful task can make it attractive outside of its associated outcome, and how contextual factors mold their perceived costs (see Inzlicht, Shenhav, and Olivola, 2018, for a review on the attractiveness of effort). Understanding these additional qualities could improve our ability to motivate completion of effortful daily activities, such as schoolwork or physical exercise.

  What is often missing from existing accounts is the common case of choosing among types of behavioral demands. Studies have traditionally compared the reduction of subjective value imposed by each type of demand, often through discounting functions that capture how much less appealing a reward becomes as a function of the amount or intensity of its associated behavior [1,3,5,8,9,14]. Although this line of work suggests that discounting functions are similar for delay and effort [8,15,16], there is ongoing debate about the shape of the discounting function across effort types [2,8,11,17,18], and discount rates do not always correlate across cost domains [9,16]. However, sporadic preferences for effort could arise from the availability of other types of demands in the same choice environment—in other words, from its context (Kool & Botvinick, 2018). For example, exerting high effort for a reward might be aversive, but the same level of work might become valuable when compared to the boredom of passive waiting (e.g. taking the cumbersome way home rather than waiting for a delayed train, even if the resulting time of arrival is the same). Conversely, even though people tend to donate more to charitable events that require them to perform effortful activities (i.e. the martyrdom effect), they are also less likely to engage in such events once an easier alternative to raise funds becomes available (Olivola and Shafir, 2013). Recent work has proposed that the subjective costs produced by delays and cognitive effort is modulated by the possibility of missing out on better concurrent alternatives, namely the opportunity cost of time (OC; Fawcett et al., 2012; Kurzban et al, 2013; Otto and Daw, 2019). This evidence sets the stage for examining what people would actually select when given the option for multiple types of demands, including the unexplored question of deciding between cognitive and physical effort (Inzlicht, Shenhav, and Olivola, 2018).

  Single choice foraging paradigms provide a natural way to probe preferences among demands, by asking individuals whether the current offer outweighs what could be obtained and experienced during that time instead (Stephens & Krebs, 2019). Foraging has recently attracted high scientific interest due to its ecological validity, which is rooted in evolutionary choice behaviors (Hayden 2018; Mobbs, Trimmer, Blumstein, & Dayan, 2018), and offers a number of desirable qualities. First, it allows for the estimation of reward maximizing choice patterns based on the opportunity cost of time (Charnov, 1975; Kacelnik and Krebs, 1977), which people can readily approximate and persistently follow (Constantino and Daw, 2015 ADD). One can then interpret observed deviations in acceptance rates from such optimal behavior in addition to comparing them among demands. Second, the sequential nature of foraging allows individuals to emphasize different elements of the decision context. For example, the cost (or value) of effort or delay could globally depend on the availability of alternative demands (as suggested above), impacting the perceived opportunity cost of time. Alternatively, it could be intrinsic and robust to such manipulation, and/or depend on task-specific factors that adapt once individuals experience different forms of behavioral costs. Foraging rests on an abundant modeling history that can help to computationally capture such emphases (Stephens & Krebs, 2019).

  In this study, we tested the hypothesis that preferences would vary depending on whether individuals face a single or multiple types of behavioral demands (i.e. cognitive effort, physical effort, and delay). We leveraged the ecological and computational richness of foraging paradigms to provide new insights on a number of open questions. First, the existence of a well-defined optimal choice strategy in foraging designs helped us identify when demands boost the value of rewards (i.e. overharvesting poor offers). Second, increasing the level of exposure to types of demands over two experiments allowed us to asses whether their cost was intrinsic or relative. Finally, confronting participants with both mental and physical effort let us probe whether these preferences were domain general or task-specific, and in the case of the latter, isolate factors that influence preferences beyond the often cited associative relationship between effort and outcome (e.g. features of an effortful task that makes it more attractive).

  Confirming our predictions, we found that individuals could approximate optimality by preferring higher rewards and shorter delays, regardless of the richness of the environment, or whether they faced a single or intermixed demands. When faced with a single type of demand, people in the cognitive effort group accepted more trials than optimal, whereas physical effort and wait groups underaccepted, partially matching our predictions and highlighting that cognitive effort can add value to outcomes. However, against our expectations, we found that a similar preference for cognitive effort disappeared over time as decision makers foraged for types of demands in addition to rewards. Computational modeling suggested that cognitive effort was valuable because it shortened the subjective elapsed time of trials, a perception that could be recalibrated as individuals experienced multiple forms of demands (a detailed description of our preregistered hypotheses can be found on https://osf.io/2rsgm/). Our results expand our understanding of the attractiveness of effort, highlighting a role for features in the decision environment that are independent from a pure association between effort and reward, and point to characteristics that can be leveraged in order to motivate effort engagement in diverse everyday scenarios.

>>>>

  Evaluating whether to pursue a desired outcome often requires assessing the demands required for obtaining it. In psychology and economics, demands for both time and effort have traditionally been identified as costs that detract from the net subjective value of rewards (Westbrook & Braver, 2015). Indeed, decision makers tend to prefer low-effort and immediate rewards over larger rewards that demand longer delays and higher effort (Ainslie, 1975; Hull, 1943; Kool & Botvinick, 2018; Kool, McGuire, Rosen, & Botvinick, 2010; Treadway, Buckholtz, Schwartzman, Lambert, & Zald, 2009; Westbrook, Kester, & Braver, 2013). However, people often change their attitudes towards exerting effort in everyday decisions, resulting in pursual of effortful avenues instead. Emerging experimental and theoretical work has chiefly explained this phenomenon by suggesting a reciprocal relationship between effort and outcome, whereby exerting effort boosts the perceived value of its outcomes (Kacelnik and Marsh, 2002; Mochon, Norton, and Ariely, 2012; Hernandez Lallement et al., 2014), and positive outcomes imbue value into their preceding effortful behavior (i.e. learned industriousness, Eisenberger, 1992). Though intuitively valid and empirically supported, this explanation leaves open questions about what intrinsic features of an effortful task can make it attractive outside in its own right, and how contextual factors mold its perceived costs (Inzlicht, Shenhav, and Olivola, 2018). Understanding these qualities could improve our ability to motivate completion of effortful daily activities, such as schoolwork or physical exercise.

  What is often missing from existing accounts is the common case of choosing among types of behavioral demands. Studies have traditionally compared the reduction of subjective value imposed by each type of demand, often through discounting functions that capture how much less appealing a reward becomes as a function of the amount or intensity of its associated behavior (Ainslie, 1975; Frederick et al., 2002; Kable and Glimcher, 2007; Kool et al., 2010; Westbrook, Kester, and Braver, 2013; Green, Fristoe, and Myerson, 1994). Although this line of work suggests that discounting functions are similar for delay and effort (Prevost et al., 2010; Massar et al., 2015; Seaman et al., 2018), there is ongoing debate about the shape of the discounting function across effort types (Arulpragasam et al., 2018; Prevost et al., 2010; Chong et al., 2016; Klein-Flugge et al., 2016; Kool and Botvinick, 2018), and discount rates do not always correlate across cost domains (Westbrook and Braver, 2013; Seaman et al., 2018). However, sporadic preferences for effort could arise from the availability of other types of demands in the same choice environment—in other words, from its context (Kool & Botvinick, 2018). For example, exerting high effort for a reward might be aversive, but the same level of work might become valuable when compared to the boredom of passive waiting (e.g. taking the cumbersome way home rather than waiting for a delayed train, even if the resulting time of arrival is the same). Conversely, even though people tend to donate more to charitable events that require them to perform effortful activities (i.e. the martyrdom effect), they are also less likely to engage in such events once an easier alternative to raise funds becomes available (Olivola and Shafir, 2013). Recent work has proposed that the subjective costs produced by delays and cognitive effort is modulated by the possibility of missing out on better concurrent alternatives, namely the opportunity cost of time (OC; Fawcett et al., 2012; Kurzban et al, 2013; Otto and Daw, 2019). This evidence sets the stage for examining what people would actually select when given the option for multiple types of demands, including the unexplored question of deciding between cognitive and physical effort (Inzlicht, Shenhav, and Olivola, 2018).

  Single choice foraging paradigms provide a natural way to probe preferences among demands, by asking individuals whether the current offer outweighs what could be obtained and experienced during that time instead (Stephens & Krebs, 2019). Foraging has recently attracted high scientific interest due to its ecological validity, which is rooted in evolutionary choice behaviors (Hayden 2018; Mobbs, Trimmer, Blumstein, & Dayan, 2018), and offers a number of desirable qualities. First, it allows for the estimation of optimal, reward maximizing choice patterns based on the opportunity cost of time (Charnov, 1975; Krebs et al., 1977). We can therefore interpret observed deviations in acceptance rates from such optimal behavior in addition to comparing them among demands. Second, the sequential nature of foraging allows individuals to emphasize different elements of the decision context. For example, the cost (or value) of effort or delay could globally depend on the availability of alternative demands (as suggested above), impacting the perceived opportunity cost of time. Alternatively, it could be intrinsic and robust to such manipulation, and/or depend on task-specific factors that are adjusted once individuals experience different forms of behavioral costs. Foraging rests on an abundant modeling history that can help to computationally capture such emphases (Stephens & Krebs, 2019).

  In this study, we tested the hypothesis that preferences would vary depending on whether individuals face a single or multiple types of behavioral demands (i.e. cognitive effort, physical effort, and delay). We leveraged the ecological and computational richness of foraging paradigms to provide new insights on a number of open questions (Inzlicht, Shenhav, and Olivola, 2018). First, the existence of a well-defined optimal choice strategy in foraging designs helped us identify when demands boost the value of rewards (i.e. overharvesting poor offers). Second, increasing the level of exposure to types of demands across two experiments allowed us to asses whether their cost was intrinsic or relative. Finally, confronting participants with both mental and physical effort let us probe whether these preferences were domain general or task-specific, and in the case of the latter, isolate factors that influence preferences beyond the often-cited associative relationship between effort and outcome.

  We found mixed support for our pre-registered hypotheses, which are detailed in https://osf.io/2rsgm/. Confirming our predictions, we found that individuals could approximate optimality by preferring higher rewards and shorter delays, regardless of the richness of the environment or whether they faced a single or intermixed demands. When faced with a single type of demand, people in the cognitive effort group accepted more trials than optimal, whereas physical effort and wait groups under-accepted, partially matching our predictions and highlighting that cognitive effort can add value to outcomes. However, against our expectations, we found that a similar preference for cognitive effort disappeared over time as decision makers foraged for types of demands in addition to rewards. Computational modeling suggested that cognitive effort was valuable because it shortened the subjective elapsed time of trials, a perception that could be recalibrated as individuals experienced multiple forms of demands. The shortened time sensitivity of cognitive effort could be due to its constant recruitment of working memory, impeding estimations of the nominal time, or because of the discrete nature of the task (versus a continuous configuration of the physical and wait tasks). These results expand our understanding of the attractiveness of effort, bringing forth outcome-independent features that can be leveraged to motivate effort engagement in diverse everyday scenarios.
